Closing the gap between statistical and scientific workflows for improved inference in ecology

### 2025-07-17 revision/response_r1.tex

{\bf Response to reviewers' comments:}

{\bf Reviewer 1}

\begin{mybox}
\emph{This opinion manuscript (MS) argues that adapting a modern statistical workflow will help resolve some key issues in forecasting models used in ecology to inform policy about biodiversity loss and climate changes impacts. Specifically, the authors identity two specific “disconnected paradigms” for making such forecasts and argue that a workflow adopted from the Bayesian literature, and that emphasizes data simulation, would help unify these paradigms and make for a better scientific understanding of these important issues. I found the MS well-written, concise and easy to read, and well-organized with clearly laid out challenges and understandable solutions. I agree with the central proposal that the adoption of a “principled” workflow would be beneficial in many areas of ecology, and these types of studies in particular. I have no experience with these specific types of ecological models, but I do have experience with the Bayesian workflow underlying their key recommendation. So, I will focus my review on the recommendation and not the challenge. Below I outline several major concerns/questions, and below that minor comments and corrections.}
\end{mybox}

We thank the reviewer for their positive comments regarding our manuscript. 
We have worked to clarify the points raised by the reviewer in this new version.

@@

\begin{mybox}
\emph{\textbf{Inadequate justification of simulation.}\\
Simulating data and fitting models is key to the step-by-step workflow but I think the MS lacks sufficient detail and motivation. The methods described come from the Bayesian literature, where simulating parameters from priors (prior predictive/pushforward) or posterior (posterior predictive or retrodictive) and then data, are natural. But a quick peek at the cited literature suggests that Bayesian models are not used. So how would one adapt them?}
\end{mybox}

Bayesian models are more and more frequently used in ecology (we cite for example \citealp{Johnson2024}). % I think this one is Bayesian right?
We clarified that some terms are specific to Bayesian inference but that data simulation still apply for all (\href{file:forecastflows_r1\#lintarget:priorPC}{line \ref*{lin:priorPC}}, \href{file:forecastflows_r1\#lintarget:PPC}{line \ref*{lin:PPC}}, \href{file:forecastflows_r1\#lintarget:broadcontext}{line \ref*{lin:broadcontext}}).

% Below, AG says not relevant
\begin{mybox}
\emph{On L76 they say “... by fixing parameters to some reasonable range of values (which is straightforward if the parameters are interpretable)” but I disagree with this. Do you suggest analysts develop priors (using, e.g., Betancourt’s cited workflow case study), and then simulate parameters from those, but ignore them in the subsequent fitting to empirical data? Or take a range for each parameter and simulate parameters across N equal steps for each one? Posterior predictive checks (retrodiction) are known to be conservative due to the double-use of the data (see Conn et al. 2018 [https://doi.org/10.1002/ecm.1314]). How does that affect your recommendations and how the analyst should use this tool?}
\end{mybox}

Thank you for this interesting point. We have work to clarify the data simulaton step (\href{file:forecastflows_r1\#lintarget:firstsim}{line \ref*{lin:firstsim}}, \href{file:forecastflows_r1\#lintarget:repeatsim}{line \ref*{lin:repeatsim}}, \href{file:forecastflows_r1\#lintarget:priorPC}{line \ref*{lin:priorPC}}), and now highlight one should develop specific posterior predictive checks for the question of interest (\href{file:forecastflows_r1\#lintarget:howtoPPC}{line \ref*{lin:howtoPPC}}). However, we believe an extended discussion of the PPC advantages and inconvenients is outside the word limit we have.

@@

{\bf Reviewer 2}

\begin{mybox}
\emph{The paper is well written and in a style that is accessible to a wide audience. The topic fits under the theme of “workflow for applied data analysis” by proposing a workflow that if used thoughtfully could lead to less divergence in forecasts and therefore more trust in the scientific process. The proposed workflow is similar to the Bayesian workflow outlined in Gelman et al. (2020) and is translated to two different approaches to ecological forecasting. I think the proposed workflow could be applied or modified to be used with a wide range of research questions within ecological forecasting if careful attention is paid to nuances and idiosyncrasies associated with each unique research question, and I think this necessary attention to question-specific nuances could be highlighted more in the paper. There are places where I feel more specificity (or clarity) with respect to these nuances could improve the manuscript and its potential for impact with respect to building trust in the scientific process. There is also a lack of specificity in how, exactly, this type of workflow could be implemented because there is no example application to data. I have outlined these areas under “Major comments” and provided minor editorial comments in “Minor Comments.”}
\end{mybox}

% Note: I am an applied statistician with experience collaborating with ecologists and developing
% statistical methodology for large-scale long-term monitoring programs. The bulk of my work
% is more concerned with explanatory than predictive inference and my knowledge of building
% ecological forecasting models for inference on a global scale is limited. I reviewed the paper
% through this lens.

We are glad that the reviewer found our manuscript well-written. 
We thank them for the feedback, which has improved the manuscript.

### 2025-07-17 revision/forecastflows_r1.tex
small modif. to figure C in box
continue working on revision

\begin{figure}[h]
	\centering
	\includegraphics{../figures/figure_modeldataspaces_revised}
	\caption{Different trajectories through model-data space during model development, from pre-model/pre-data to post-model/post-data quadrants.
	Panel A serves as a legend, showing where model development should begin (carefully thinking about the hypotheses we want to test) and when inference should be made, for either explanatory purpose (estimation) or forecasting (prediction).
	Panel B illustrates that in current trend estimation approaches, most time is spent in the post-model/post-data quadrant, where models are directly fitted to empirical data---often with no prior evaluation of the model.
	Panel C illustrates the typical process-based model development. A lot of time is spent building and calibrating (fitting) separate submodels, without clear distinction between model building and model calibration and without evaluating the full model structure a priori. Additionally, the specific hypotheses the model is designed to test are rarely clearly stated.
	Panel D illustrates the workflow highlighted in this paper (see main text and Figure 2 for details). Ideally, it begins in the pre-model/pre-data quadrant, where the research question is defined. The next step moves to the post-model/pre-data quadrant, where a model is built---independently of the data. This is where we should spend a good deal of time evaluating model behavior using simulated data. Once we are satisfied with our model, we can move to the post-model/post-data quadrant, where the model is fitted to real data. Then, we perform retrodictive checks to compare model predictions to observations, which likely give us some feedback to refine our model.}
	\label{fig:modeldata}
\end{figure}

@@

\begin{figure}[h]
	\centering
	\vspace*{-0.1cm}
    \hspace*{-1.5cm}
	\includegraphics{../figures/figure_worflow_wsteps}
	\caption{The workflow we propose here \citep[which builds from recent advances in workflows,][]{betanworkflow,Gelman2020,Schad2020,grinsztajn2021,vandeschoot2021,Wolkovich2024} focuses on iterative feedbacks between the research question, model and data. Most research should start at Step 1, with the research question, followed by extensive model building and evaluation using simulated data (Steps 2 and 3) before proceeding to fitting the model on empirical data and examining it through retrodictive checks (Steps 4 and 5). Within this workflow, forecasting is a natural output of the process and not a separate process or one available for only certain modeling approaches.  The integration of observations (the data stream) occurs at Step 4---only after the model has been thoroughly evaluated---and can also highlight opportunities to collect new data and enrich the data stream.}
	\vspace*{-0.5cm}
	\label{fig:workflow}
\end{figure}

@@

\centerline{\bf A common workflow to bridge trend estimation and forecasting} % jdavies15april: would it be possible to label this panel with components of the LPI to provide a clearer illustration of how the framework is implemented? % [FLAG: not sure I understood]
\vspace*{-3mm}
{\begin{multicols}{2}
\begin{minipage}[t]{\linewidth}
	\vfill
\includegraphics[width=\linewidth]{../figures/iterativeworkflow_details}
\vfill
% \vspace{-0.5cm}\captionsetup{font=footnotesize}\captionof{figure}{Caption}\label{fig:trends}
\vspace*{3mm}
\end{minipage}

>>>>

\begin{figure}[h]
	\centering
	\includegraphics{../figures/figure_modeldataspaces_revised}
	\caption{Different trajectories through model-data space during model development, from pre-model/pre-data to post-model/post-data quadrants.
	Panel A serves as a legend, showing where model development should begin (carefully thinking about the hypotheses we want to test) and when inference should be made, for either explanatory purpose (estimation) or forecasting (prediction).
	Panel B illustrates that in current trend estimation approaches, most time is spent in the post-model/post-data quadrant, where models are directly fitted to empirical data---often with no prior evaluation of the model.
	Panel C illustrates the typical process-based model development. A lot of time is spent building and calibrating (fitting) separate submodels (represented by the separated ovals), without clear distinction between model building and model calibration and without evaluating the full model structure a priori. Additionally, the specific hypotheses the model is designed to test are rarely clearly stated.
	Panel D illustrates the workflow highlighted in this paper (see main text and Figure 2 for details). Ideally, it begins in the pre-model/pre-data quadrant, where the research question is defined. The next step moves to the post-model/pre-data quadrant, where a model is built---independently of the data. This is where we should spend a good deal of time evaluating model behavior using simulated data. Once we are satisfied with our model, we can move to the post-model/post-data quadrant, where the model is fitted to real data. Then, we perform retrodictive checks to compare model predictions to observations, which likely give us some feedback to refine our model.}
	\label{fig:modeldata}
\end{figure}

@@

\begin{figure}[h]
	\centering
	\vspace*{-0.1cm}
    \hspace*{-1.5cm}
	\includegraphics{../figures/figure_worflow_wsteps}
	\caption{The workflow we propose here \citep[which builds from recent advances in workflows,][]{betanworkflow,Gelman2020,Schad2020,grinsztajn2021,vandeschoot2021,Wolkovich2024} focuses on iterative feedbacks between the research question, model and data. Most research should start at Step 1, with the research question, followed by extensive model building and evaluation using simulated data (Steps 2 and 3) before proceeding to fitting the model on empirical data and examining it through retrodictive checks (Steps 4 and 5). Within this workflow, forecasting is a natural output of the process and not a separate process or one available for only certain modeling approaches.  The integration of observations (the data stream) occurs at Step 4---only after the model has been thoroughly evaluated---and can also highlight opportunities to collect new data and enrich the data stream. This figure is expanded from Figure \ref{fig:modeldata}d.}
	\vspace*{-0.5cm}
	\label{fig:workflow}
\end{figure}

@@

\centerline{\bf A common workflow to bridge trend estimation and forecasting} % jdavies15april: would it be possible to label this panel with components of the LPI to provide a clearer illustration of how the framework is implemented? % [FLAG: not sure I understood]
\vspace*{-3mm}
{\begin{multicols}{2}
\begin{minipage}[t]{\linewidth}
	\vfill
\includegraphics[width=\linewidth]{../figures/iterativeworkflow_details_revised}
\vfill
% \vspace{-0.5cm}\captionsetup{font=footnotesize}\captionof{figure}{Caption}\label{fig:trends}
\vspace*{3mm}
\end{minipage}
